One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-26 paper/model.tex
final draft

\section{Linear density waves}\label{wkb}
In this paper we will mainly use non-linear hydrodynamic simulations to investigate the
stability of disc equilibra described 
above. However, it is useful to  interpret simulation results
using linear density theory. We list here the main definitions and
results from 2D linear theory, taken from \cite{lin11b} and
\cite{shu91}.   In particular, we highlight a key feature of
locally isothermal discs that enables angular momentum exchange
between linear perturbations and the background disc.

@@

The important consequence of the background torque is the possibilty
of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
(negative) and the background torque $\tbg$ is positive (negative),
then the local angular momentum density of the linear disturbance
further increases (decreases). This implies the amplitude of the disturbance
may grow by exchanging angular momentum with the background
disc. In our disc models with $c_s^2\propto R^{-q}$, we expect the
growth rate of such an instability to scale with $q$.

@@

\subsubsection{Evolution of non-axisymmetirc modes}
The disc evolution is quantified using mode amplitudes and angular
momenta as follows. We list the 2D definitions with obvious 3D generalisations. 
A hydrodynamic variable $f$ (e.g. $\Sigma$) is written as 
\begin{align}
  f(R,\phi,t) &= \sum_{m=-\infty}^{\infty}f_m(R,t)\exp{\ii m \phi} \notag\\
  &= f_0 + 2 \real\left[\sum_{m=1}^\infty f_m \exp{(\ii
      m\phi)}\right], 
\end{align}
where the $f_m$ may be obtained from Fourier transform in $\phi$.

>>>>

\section{Linear density waves}\label{wkb}
In this paper we will mainly use non-linear hydrodynamic simulations to investigate the
stability of disc equilibria described 
above. However, it is useful to  interpret simulation results
using linear density theory. We list here the main definitions and
results from 2D linear theory, taken from \cite{lin11b} and
\cite{shu91}.   In particular, we highlight a key feature of
locally isothermal discs that enables angular momentum exchange
between linear perturbations and the background disc.

@@

The important consequence of the background torque is the possibility
of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
(negative) and the background torque $\tbg$ is positive (negative),
then the local angular momentum density of the linear disturbance
further increases (decreases). This implies the amplitude of the disturbance
may grow by exchanging angular momentum with the background
disc. In our disc models with $c_s^2\propto R^{-q}$, we expect the
growth rate of such an instability to scale with $q$.

@@

\subsubsection{Evolution of non-axisymmetric modes}
The disc evolution is quantified using mode amplitudes and angular
momenta as follows. We list the 2D definitions with obvious 3D generalisations. 
A hydrodynamic variable $f$ (e.g. $\Sigma$) is written as 
\begin{align}
  f(R,\phi,t) &= \sum_{m=-\infty}^{\infty}f_m(R,t)\exp{\ii m \phi} \notag\\
  &= f_0 + 2 \real\left[\sum_{m=1}^\infty f_m \exp{(\ii
      m\phi)}\right], 
\end{align}
where the $f_m$ may be obtained from Fourier transform in $\phi$.

### 2014-11-26 paper/paper.tex
final draft

\section{Introduction}\label{intro}
% Protoplanetary discs are expected to contain complex structures
% \citep{armitage11}, which directly impact planet formation theories.  
An exciting development in the study of circumstellar 
discs is the direct observation of large-scale, non-axisymmmetric
structures within them. These
include lopsided dust distributions 
\citep{marel13,fukagawa13,casassus13,isella13,perez14,follette14} and
spiral arms 
\citep{hashimoto11,muto12,boccaletti14,grady13,christiaens14,avenhaus14}.

@@

This paper is organised as follows. In \S\ref{model} we list the
governing equations and describe the disc equilibria under
consideration. \S\ref{methods} describes the hydrodynamic codes
employed in this work and several diagnostic measures. We briefly
review basic results from linear theory in \S\ref{wkb} to aid our
interpretation of numerical simulations.  Our main results are presented in 
\S\ref{results2d} for 2D simulations and \S\ref{results3d} for 3D
simulations, and we further analyse them in \S\ref{discussions}. 
We summarise in \S\ref{summary}, with a discussion on
speculations for future work.

>>>>

\section{Introduction}\label{intro}
% Protoplanetary discs are expected to contain complex structures
% \citep{armitage11}, which directly impact planet formation theories.  
An exciting development in the study of circumstellar 
discs is the direct observation of large-scale, non-axisymmetric
structures within them. These
include lopsided dust distributions 
\citep{marel13,fukagawa13,casassus13,isella13,perez14,follette14} and
spiral arms 
\citep{hashimoto11,muto12,boccaletti14,grady13,christiaens14,avenhaus14}.

@@

This paper is organised as follows. In \S\ref{model} we list the
governing equations and describe the disc equilibria under
consideration. We briefly review basic results from linear theory in
\S\ref{wkb} to aid our interpretation of numerical simulations. 
\S\ref{methods} describes the hydrodynamic codes
employed in this work and several diagnostic measures. Our main results are presented in 
\S\ref{results2d} for 2D simulations and \S\ref{results3d} for 3D
simulations, and we further analyse them in \S\ref{discussions}. 
We summarise in \S\ref{summary}, with a discussion on
speculations for future work.

### 2014-11-26 paper/results.tex
final draft

\subsection{Dependence on the imposed temperature profile}
The simulations above suggest the growth of the $m=1$ spiral mode
can be attributed to the background torque applied on linear pertubrations
as a result of an imposed temperature gradient, as discussed in
\S\ref{global_cons}. According to Eq. \ref{baroclinic_torque}, this
torque is proportional to $q$, the power-law index used to set the
sound-speed profile (Eq. \ref{sound-speed}) in our simulations. 
We confirm this through a series of simulations with $q\in[0,1]$, 
initialised with $m=1$ perturbations. However, to keep the Toomre $Q$
profiles similar, we adjust the surface density power-law index such
that $s = (3+q)/2$.

>>>>

\subsection{Dependence on the imposed temperature profile}
The simulations above suggest the growth of the $m=1$ spiral mode
can be attributed to the background torque applied on linear perturbations
as a result of an imposed temperature gradient, as discussed in
\S\ref{global_cons}. According to Eq. \ref{baroclinic_torque}, this
torque is proportional to $q$, the power-law index used to set the
sound-speed profile (Eq. \ref{sound-speed}) in our simulations. 
We confirm this through a series of simulations with $q\in[0,1]$, 
initialised with $m=1$ perturbations. However, to keep the Toomre $Q$
profiles similar, we adjust the surface density power-law index such
that $s = (3+q)/2$.

### 2014-11-26 paper/summary.tex
final draft

\section{Summary and conclusions}\label{summary}
In this paper, we have carried out direct numerical hydrodynamic
simulations of radially structured, self-gravitating locally
isothermal discs.     
We find such systems can be unstable to low-frequency perturbations
with azimuthal wavenumber $m=1$, which lead to the development of one-armed 
trailing spirals that persist for at least $O(10^2)$ orbits. The 
spiral pattern speed is smaller than the local disc rotation, and
growth rates are $O(10^{-2}\Omega)$ which gives a characteristic
growth time of $O(10)$ orbits.   
%While this is slow
%compared to the local rotation, it is dynamical at co-rotation.

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to verify the growth of  
one-armed spirals in our disc models is due to the imposed temperature
profile and disc structure. The spiral is confined between two
$Q$-barriers owing to a surface density bump. Growth rates increased
with the magnitude of the imposed temperature gradient, and one-armed
spirals did not develop in strictly isothermal simulations.  
We find the instability behaves similarly in 2D and 3D. However, in 3D
the spiral disturbance can become radially global away from 
the midplane.

@@

% prelim sims show they are long lasting 
% balancing heating and cooling 
% should not get fragmentation (CR away from spiral) 
In the deeply non-linear regime, the one-armed spiral may 
shock and deposit negative angular momentum onto
the background disc. The spiral amplitude would saturate by gaining
positive angular momentum. However, if the temperature gradient is
maintained, it may be possible to achieve a balance between the gain
of negative angular momentum through the temperature gradient, and the
gain of positive angular momentum through shock dissipation. We remark  
that fragmentation is unlikely because the co-rotation radius 
is outside the bulk of the spiral arm, in the non-self-gravitating
portion of the disc \citep{durisen08}. In order
to study these possibilities, improved numerical models are needed to
ensure total angular momentum conservation on timescales much longer
than that considered in this paper.

@@

>>>>

\section{Summary and conclusions}\label{summary}
In this paper, we have carried out direct numerical hydrodynamic
simulations of radially structured, self-gravitating locally
isothermal discs.     
We find such systems can be unstable to low-frequency perturbations
with azimuthal wavenumber $m=1$, which leads to the development of an one-armed 
trailing spiral that persist for at least $O(10^2)$ orbits. The spiral
pattern speed is smaller than the local disc rotation and
growth rates are $O(10^{-2}\Omega)$ which gives a characteristic
growth time of $O(10)$ orbits.   
%While this is slow
%compared to the local rotation, it is dynamical at co-rotation.

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to show that the growth of  
one-armed spirals in our disc models is due to the imposed temperature 
profile and disc structure. Growth rates increased 
with the magnitude of the imposed temperature gradient, and one-armed
spirals did not develop in strictly isothermal simulations. The spiral
is confined between two $Q$-barriers owing to a surface density bump. 
We find the instability behaves similarly in 2D and 3D. However, in 3D
the spiral disturbance can become radially global away from 
the midplane.

@@

% prelim sims show they are long lasting 
% balancing heating and cooling 
% should not get fragmentation (CR away from spiral) 
In the deeply non-linear regime, the one-armed spiral may 
shock and deposit negative angular momentum onto
the background disc. The spiral amplitude would saturate by gaining
positive angular momentum. However, if the temperature gradient is
maintained, it may be possible to achieve a balance between the gain
of negative angular momentum through the background torque, and the
gain of positive angular momentum through shock dissipation. We remark  
that fragmentation is unlikely because the co-rotation radius 
is outside the bulk of the spiral arm, in the non-self-gravitating
portion of the disc \citep{durisen08}. In order
to study these possibilities, improved numerical models are needed to
ensure total angular momentum conservation on timescales much longer
than that considered in this paper.

@@

%theoretical investigation on effect on other m modes
